---
title: Hispanics in New Mexico
subtitle: A demographic study with R and TidyCensus
description: New Mexico has a Latino population unlike anywhere else in the US. This article shows how to use R and tidycensus to explore the data.
output: 
  html_document:
    toc: true
    keep_md: true
    self_contained: true
date: 2025-05-24
code-overflow: wrap
categories: [R, Data Science, GIS]
draft: false
---

```{r setup, include=FALSE, paged.print=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Motivation

```{r message=FALSE}
library(tidycensus)
library(tidyverse)
library(tigris)
library(patchwork)
library(sf)
options(tigris_use_cache = TRUE)
```

In a [prior article](../nm-hispanic-demographics/index.qmd), I used R and TidyCensus to perform a spatial analysis of the various Hispanic groups in New Mexico. We saw that nearly half of New Mexicans are Hispanic, a much larger proportion than any other state, and that they are largely split into two groups, not including the unclear "Hispanic-White" designation. The Mexican Hispanic population is more concentrated along the southern border, while the Spanish Hispanics, the Hispanos who were in New Mexico when it joined the United States, and identify with their European origins rather than their Mexican origins, are concentrated in the northeast. These are descendents of some of the earliest settlers in North America.

Despite the geographic seperation state-wide, more people of both groups live in Bernalillo County, where the largest city in the state, Albuquerque, is located. While the separation across the state seems perfectly nature given the history, does such a division exist in an urban setting where large numbers of both groups interact in close proximity.

# Selecting datasets

The [US Census](https://www.census.gov/) website provides many views of their data. The 5-year [American Community Survey](https://en.wikipedia.org/wiki/American_Community_Survey), which we have been using, provides many other "geographies" than state and county. One which we will use here is `tract`, which provides data for each census tract, the basic geographical "unit" in the survey. For this analysis, I want to look at the greater Albuquerque area, which includes the sister city of Rio Rancho, and surrounding areas. As a point of comparison, I'd like to consider El Paso. Geographically, it is more naturally attached to New Mexico, and its sister city is Las Cruces, which is in New Mexico.

Other potential sets include *Combined Statistical Areas*(CSAs) and *Core Based Statistical Areas*(CBSAs). CBSAs are defined by the census bureau, and based on geographic proximity to a given metropolitan center, taking into account interactions between contiguous regions. These areas may extend outside of the county in which the metropolitan center is located. The CSAs, on the other hand, group together different nearby cities.

Since I'm interested in "Greater Albuquerque", either of these are possibilities. I'll start by comparing the two geographically, to see which looks best.

```{r}
#| eval: false


v23 <- load_variables(year = 2023, "acs5/profile", cache = TRUE)
View(v23)
```

I'll start by downloading the tract-level geographies using `tigris::tracts()`. Since I intend to look and El Paso and Las Cruces, I'll go ahead and download the data for both New Mexico and Texas. I will use an apprpriate [*Coordinate Reference System*](https://en.wikipedia.org/wiki/Coordinate_reference_system) *(CRS)* for New Mexico. (When in doubt, call `crsuggest::suggest_crs()` on the data frame.)

```{r}
tx_nm_tracts <- map_dfr(c("TX", "NM"), ~{
  tracts(.x, cb = TRUE, year = 2023)
}) |> 
  st_transform(32113)
```

I'll start with the Combined Statistical Areas.

```{r}
abq_csa <- combined_statistical_areas(
  cb = TRUE, year = 2023
) |> 
  filter(CSAFP == "105") |> 
  st_transform(32113)

el_paso_csa <- combined_statistical_areas(
  cb = TRUE, year = 2023
) |> 
  filter(CSAFP == "238") |> 
  st_transform(32113)

```

```{r}
places <- read_sf("data/places/tl_2023_35_place.shp")
cities <- c("Albuquerque", "Rio Rancho", "Las Cruces", 
            "Santa Fe", "Los Alamos")

city_ep <- read_sf("data/tl_2023_48_place.shp") |> 
  filter(NAME == "El Paso") |> 
  select(City = NAME, geometry) |> 
  st_as_sf() |> 
  st_centroid()

city_pts <- places |> 
  filter(NAME %in% cities) |> 
  select(City = NAME, geometry) |> 
  st_as_sf() |> 
  st_centroid()
```

```{r}
el_paso_csa_tracts <- tx_nm_tracts[el_paso_csa, ] |> 
  st_filter(el_paso_csa, .predicate = st_within)

p1 <- ggplot() +
  geom_sf(data = el_paso_csa_tracts, fill = "white", color = "grey80") +
  geom_sf(data = el_paso_csa, fill = NA, color = "cyan") +
  geom_sf_text(data = city_pts |> filter(City == "Las Cruces"), aes(label = City),
               size = 5, color = "black",
               vjust = "top",hjust = "left",
               nudge_y = -.1) +
  geom_sf_text(data = city_ep, aes(label = City),
               size = 5, color = "black",
               vjust = "top",hjust = "left",
               nudge_y = -.1) +
  labs(title = "El Paso CSA") +
  theme_void()
```

```{r}
abq_csa_tracts <- tx_nm_tracts[abq_csa, ] |> 
  st_filter(abq_csa, .predicate = st_within)

p2 <- ggplot() +
  geom_sf(data = abq_csa_tracts, color = "grey80", fill = NA) +
  geom_sf(data = abq_csa, fill = NA, color = "cyan") +
  geom_sf_text(data = city_pts |> filter(City != "Las Cruces"), aes(label = City),
               size = 5, color = "black",
               vjust = "top",hjust = "left",
               nudge_y = -.1) +
  labs(
    title = "Albuquerque CSA"
  ) +
  theme_void()
```

```{r}
p1 + p2
```

```{r}
nm_income <- get_acs(
  geography = "tract", 
  state = "NM", 
  variables = "B19013_001", 
  year = 2020, 
  geometry = T
)
```

```         
## Getting data from the 2016-2020 5-year ACS
```

```{r}
nm_income
```

## Basic plotting

``` r
plot(nm_income['estimate'])
```

## ggplot2

### Cholopleth

``` r
us_median_age <- get_acs(
  geography = "state", 
  variables = "B01002_001",
  year = 2019,
  survey = "acs1",
  geometry = T,
  resolution = "20m",
) |> 
  shift_geometry()
```

``` r
plot(us_median_age$geometry)
```

``` r
ggplot(us_median_age, aes(fill = estimate)) +
  geom_sf() +
  labs(title = "Median Age by State, 2019", 
       caption = "Data Source US Census Bureau 1-year ACS", 
       fill = "ACS estimate") +
  scale_fill_distiller(palette = "RdPu", 
                       direction = 1) +
  theme_void()
```

## tmap

``` r
bernalillo_race <- get_decennial(
  geography = "tract", 
  state = "NM", 
  county = "Bernalillo",
  variables = c(
    Hispanic = "P2_002N",
    White = "P2_005N",
    Black = "P2_006N",
    Native = "P2_007N",
    Asian = "P2_008N"    
  ),
  summary_var = "P2_001N",
  year = 2020,
  geometry = T
) |> 
  mutate(percent = 100 * value / summary_value)
```

``` r
bernalillo_race
```

### Chorpleth with tmap

``` r
library(tmap)
bernalillo_hisp <- bernalillo_race |> 
  filter(variable == "Hispanic")
bernalillo_hisp
```

``` r
tm_shape(bernalillo_hisp) +
  tm_polygons(fill = "percent")
```

``` r
hist(bernalillo_hisp$percent)
```

``` r
bernalillo_asian <- bernalillo_race |> 
  filter(variable == "Asian")

tm_shape(bernalillo_asian) +
  tm_polygons(fill = "percent")
```

``` r
hist(bernalillo_asian$percent)
```

``` r
tm_shape(bernalillo_asian) +
  tm_polygons(fill = "percent", 
              style = "quantile", 
              n = 5, 
              palette = "Greens", 
              title = "2020 US Census") +
  tm_layout(title = "Percent Asian\n by Census Tract",
            frame = F, legend.outside = T)
```

``` r
tm_shape(bernalillo_asian) + 
  tm_polygons(col = "percent",
          style = "jenks",
          n = 5,
          palette = "Purples",
          title = "2020 US Census",
          legend.hist = TRUE) + 
  tm_layout(title = "Percent Asian\nby Census tract",
            frame = FALSE,
            legend.outside = TRUE,
            bg.color = "grey70",
            legend.hist.width = 5,
            fontfamily = "Verdana")
```

## Reference elements

``` r
library(mapboxapi)
```

``` r
readRenviron("~/.Renviron")
```

``` r
bernalillo_tiles <- get_static_tiles(
  location = bernalillo_hisp, 
  zoom = 10,
  style_id = "light-v9",
  username = "mapbox"
)
```

``` r
tm_shape(bernalillo_tiles) + 
  tm_rgb() + 
  tm_shape(bernalillo_hisp) + 
  tm_polygons(col = "percent",
          style = "jenks",
          n = 5,
          palette = "Greens",
          title = "2020 US Census",
          alpha = 0.5) +
  tm_layout(title = "Percent Hispanic\nby Census tract",
            legend.outside = TRUE,
            fontfamily = "Verdana") + 
  tm_scale_bar(position = c("left", "bottom")) + 
  tm_compass(position = c("right", "top")) + 
  tm_credits("(c) Mapbox, OSM    ", 
             bg.color = "white",
             position = c("RIGHT", "BOTTOM"))
```

## Graduated symbols

``` r
tm_shape(bernalillo_hisp) + 
  tm_polygons() + 
  tm_bubbles(size = "value", alpha = 0.5, 
             col = "navy",
             title.size = "Hispanic - 2020 US Census") + 
  tm_layout(legend.outside = TRUE,
            legend.outside.position = "bottom")
```

## Faceted maps

``` r
tm_shape(bernalillo_race) + 
  tm_facets(by = "variable", scale.factor = 4) + 
  tm_polygons(fill = "percent",
          fill.scale = tm_scale_intervals(
            style = "jenks",
            #pal = "brewer.blues",
            n = 6
          ),
          #n = 6,
          #palette = "brewer.blues",
          fill.legend = tm_legend("Percent (2020 US Census)")) + 
  tm_layout(bg.color = "grey", 
            panel.label.bg.color = "white")
```

``` r
bernalillo_race
```

## Dot-density maps

``` r
bernalillo_dots <- bernalillo_race |> 
  as_dot_density(
    value = "value", 
    values_per_dot = 100, 
    group = "variable"
  )
```

``` r
background_tracts <- filter(bernalillo_race, variable == "Hispanic")

tm_shape(background_tracts) +
  tm_polygons(col = "grey", 
              fill.scale = tm_scale(
                
              ))  +
  tm_shape(bernalillo_dots) +
  tm_dots(col = "variable", palette = "Set1", 
          size = 0.005, title = "1 dot = 100 people") +
  tm_layout(legend.outside = T) +
  tm_title("Race/ethnicity\n2020 Census")
```

```{r}
#| eval: false


v23 <- load_variables(2023, "acs5", cache = TRUE)
View(v23)
```

```{r}
pob_vars = c(
  Native = "B05002_002",
  ForeignBorn = "B05002_013"
)
```

```{r}
library(tigris)
states <- states() |> 
  select(GEOID, state = NAME, geometry) |> 
  shift_geometry()
nm_counties <- counties("NM") |> 
  select(GEOID, county = NAME, geometry)
```

```{r}
immigration_23 <- get_acs(
  geography = "county",
  state = "NM",
  variables = pob_vars,
  summary_var = "B05002_001",
  year = 2023,
  cache_table = TRUE,
) |> 
  left_join(nm_counties,  by = "GEOID") |> 
  select(variable, estimate, moe,
         county, summary_est, geometry) |> 
  st_as_sf()

```

### Code

```{r}
immigration_us <- get_acs(
  geography = "state",
  variables = pob_vars,
  summary_var = "B05002_001",
  year = 2023,
  cache_table = TRUE,
) |> 
  left_join(states,  by = "GEOID") |> 
  select(variable, estimate, moe,
         state, summary_est, geometry) |> 
  st_as_sf()
```

Percentages can be more interesting than raw numbers, so I will calculate them. R makes it so easy to create new columns based on existing ones.

```{r}
immigration_23$pct_foreign = immigration_23$estimate / immigration_23$summary_est
immigration_us$pct_foreign = immigration_us$estimate / immigration_us$summary_est
```

Now we can see some basic stats:

```{r}
pct_fb_nm <- with(
  immigration_23, 
  mean(pct_foreign[variable == "ForeignBorn"]))
pct_fb_us <- with(
  immigration_us, 
  mean(pct_foreign[variable == "ForeignBorn"]))
data.frame(
  Area = c("New Mexico", "US"),
  Average = c(pct_fb_nm, pct_fb_us)
)
```

Despite having a the largest Hispanic population in the country, New Mexico has a much smaller percentage of immigrants than the overall average.

### Code

```{r}
orig_vars_23 = c(
  Europe = "B05006_002",
  Asia = "B05006_047",
  Eastern_Asia = "B05006_048",
  China = "B05006_049",
  Asia_SC = "B05006_056",
  Asia_SE = "B05006_068",
  Vietnam = "B05006_077",
  Asia_W = "B05006_079",
  Africa = "B05006_095",
  Oceania = "B05006_130",
  Latin_America = "B05006_139",
  Caribbean = "B05006_140",
  Central_America = "B05006_154",
  Mexico = "B05006_160",
  South_America = "B05006_164",
  Canada = "B05006_177"
)
```

```{r}
foreign_origin_nm <- get_acs(
  geography = "county",
  state = "NM",
  variables = orig_vars_23,
  summary_var = "B05006_001",
  year = 2023,
  cache_table = TRUE
) |> 
  left_join(nm_counties, by = "GEOID") |> 
  select(variable, estimate, moe,
            county, total = summary_est, 
            geometry) |> 
  st_as_sf()
```

### Code

```{r}
continents <- c(
  "Europe",
  "Central_America",
  "Asia",
  "South_America",
  "Africa",
  "Oceania",
  "Canada"
)
```

```{r}
counties_top_8 <- foreign_origin_nm |> 
  group_by(county) |> 
  summarise(est_sum = sum(estimate)) |> 
  slice_max(order_by = est_sum, n = 8) |>
  ungroup()
```

```{r}
#| echo: false

orig_vars_19 = c( Europe = "B05006_002", Asia = "B05006_047", Eastern_Asia = "B05006_048", China = "B05006_049", Asia_SC = "B05006_056", Asia_SE = "B05006_067", Vietnam = "B05006_076", Asia_W = "B05006_078", Africa = "B05006_091", Oceania = "B05006_117", Latin_America = "B05006_124", Caribbean = "B05006_125", Central_America = "B05006_138", Mexico = "B05006_139", South_America = "B05006_148", Canada = "B05006_161" )
```

# Hispanics

Let's look at where most Latinos live in the country.

### Code

```{r}
#| warning: false

us_hisp_pct <- get_acs(
  geography = "state",
  variables = c("B03001_002", "B03001_003"),
  summary_var = "B03001_001",
  year = 2023,
  cache_table = TRUE
) |>
  select(-moe, -summary_moe) |> 
  pivot_wider(names_from = variable, 
              values_from = estimate) |> 
  rename(State = "NAME", 
         Total = "summary_est", 
         Non_Hispanic = "B03001_002", 
         Hispanic = "B03001_003") |> 
  mutate(Hispanic_Pct = Hispanic / Total) |> 
  left_join(states, by = "GEOID") |> 
  st_as_sf() |> 
  shift_geometry()
```

```{r}
nm_hisp_pct <- get_acs(
  geography = "county",
  state = "NM",
  variables = c("B03001_002", "B03001_003"),
  summary_var = "B03001_001",
  year = 2023,
  cache_table = TRUE) |> 
  left_join(nm_counties,
            by = "GEOID") |> 
  select(-moe, -summary_moe) |> 
  pivot_wider(names_from = variable, values_from = estimate) |> 
  rename(County = "NAME", Total = "summary_est", 
         Non_Hispanic = "B03001_002", Hispanic = "B03001_003") |> 
  mutate(Hispanic_Pct = Hispanic / Total) |> 
  st_as_sf()
```

### Code

```{r}
hisp_vars = c(
  Non_Hispanic = "B03001_002",
  Hispanic = "B03001_003",
  Hispanic_Mexican  = "B03001_004",
  Hispanic_PR = "B03001_005", 
  Hispanic_Cuba = "B03001_006", 
  Hispanic_DR = "B03001_007", 
  Hispanic_CA = "B03001_008", 
  Hispanic_SA = "B03001_016", 
  Hispanic_Spaniard = "B03001_028",
  Hispanic_Spanish = "B03001_029",
  Hispanic_Spanish_American = "B03001_030",
  Hispanic_White = "B03002_013"
)

```

```{r}

hispanic_nm <- get_acs(
  geography = "county",
  state = "NM",
  variables = hisp_vars,
  summary_var = "B03001_001",
  year = 2023,
  cache_table = TRUE) |> 
  left_join(nm_counties, by = "GEOID") |> 
  select(GEOID, variable, estimate, moe,
            county, summary_est, 
            geometry) |> 
  st_as_sf() |> 
  pivot_wider(names_from = variable, values_from = estimate) |> 
  rowwise() |> 
  mutate(Hispanic_Spain = sum(c_across(Hispanic_Spaniard:Hispanic_Spanish_American), na.rm = TRUE)) |> 
  select(-c("Hispanic_Spaniard", "Hispanic_Spanish", "Hispanic_Spanish_American")) |> 
  pivot_longer(cols = Non_Hispanic:Hispanic_Spain, names_to = "variable", values_to = "estimate", values_drop_na = T) |> 
  group_by(county, variable) |> 
  mutate(pop = sum(estimate)) |> 
  distinct(county, variable, .keep_all = TRUE)
```

```{r}
hispanic_us <- get_acs(
  geography = "state",
  variables = hisp_vars,
  summary_var = "B03001_001",
  year = 2023,
  cache_table = TRUE) |> 
  left_join(states, by = "GEOID") |> 
  select(GEOID, variable, estimate, moe,
            state, summary_est, 
            geometry) |> 
  st_as_sf()
```

### Code

```{r}
hispanic_us <- get_acs(
  geography = "state",
  variables = hisp_vars,
  summary_var = "B03001_001",
  year = 2023,
  cache_table = TRUE) |> 
  rename(state = NAME) |> 
  pivot_wider(names_from = variable, values_from = estimate) |> 
  rowwise() |> 
  mutate(Hispanic_Spain = sum(c_across(Hispanic_Spaniard:Hispanic_Spanish_American), na.rm = TRUE)) |> 
  select(-c("Hispanic_Spaniard", "Hispanic_Spanish", "Hispanic_Spanish_American")) |> 
  pivot_longer(cols = Non_Hispanic:Hispanic_Spain, names_to = "variable", values_to = "estimate", values_drop_na = T)|> 
  group_by(state, variable) |> 
  mutate(pop = sum(estimate)) |> 
  distinct(variable, .keep_all = TRUE) |> 
  mutate(percent = pop / summary_est) |> 
  left_join(states, by = "GEOID") |> 
  select(variable, pop, summary_est,
         state = state.y, geometry) |> 
  st_as_sf()
```

```{r}
hispanic_us$percent = hispanic_us$pop / hispanic_us$summary_est
```

```{{r}}
hispanic_us <- get_acs(
  geography = "state",
  variables = hisp_vars,
  summary_var = "B03001_001",
  year = 2023,
  cache_table = TRUE) |>
  rename(state = NAME) |> 
  pivot_wider(names_from = variable, values_from = estimate) |> 
  rowwise() |> 
  mutate(Hispanic_Spain = sum(c_across(Hispanic_Spaniard:Hispanic_Spanish_American), na.rm = TRUE)) |> 
  select(-c("Hispanic_Spaniard", "Hispanic_Spanish", "Hispanic_Spanish_American")) |> 
  pivot_longer(cols = Non_Hispanic:Hispanic_Spain, names_to = "variable", values_to = "estimate", values_drop_na = T)|> 
  group_by(state, variable) |> 
  mutate(pop = sum(estimate)) |> 
  distinct(variable, .keep_all = TRUE) |> 
  mutate(percent = pop / summary_est) |> 
  left_join(states, by = "GEOID") |> 
  select(variable, pop, summary_est,
         state = state.y, geometry) |> 
  st_as_sf()
```

```{r}
hispanic_nm$percent <- hispanic_nm$pop / hispanic_nm$summary_est
```
