---
title: Exporing New Mexico Demographics with TidyCensus
subtitle: A subtitle
description: Some description
output: 
  html_document:
    toc: true
    keep_md: true
    self_contained: true
date: 2025-05-24
categories: [R, TidyCensus, Data Science]
draft: false
---

```{r setup, include=FALSE, paged.print=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This article will explore some interesting aspects of New Mexico's racial demographics using R and the TidyCensus package. The US Census Bureau makes an amazing amount of data available to the public on <https://census.gov>. The site allows you to easily search for and download demographic information with myriad variables and geographic subsets. The site also provides an API for programmatic access, and R is fortunate to have the TidyCensus package to easily grab data of interest to the programmer.

New Mexico has relatively large Native American and Hispanic populations. Nearly half of New Mexico residents are Hispanic, compared to a national average of around 20% (we'll calculate it later). The profile of Latinos in New Mexico is very different from the rest of the country, as we shall see.

## Getting started

::: {.callout-tip collapse=true icon=false}
## Code

```{r message=FALSE}
library(tidycensus)
library(tidyverse)
library(tigris)
library(patchwork)
library(sf)
options(tigris_use_cache = TRUE)
```

```{r}
states <- states() |> 
  select(GEOID, state = NAME, geometry)
nm_counties <- counties("NM") |> 
    select(GEOID, county = NAME, geometry)
```

```{r}
us_hisp_pct <- get_acs(
  geography = "state",
  variables = c("B03001_002", "B03001_003"),
  summary_var = "B03001_001",
  year = 2023,
  cache_table = TRUE
) |>
  select(-moe, -summary_moe) |> 
  pivot_wider(names_from = variable, values_from = estimate) |> 
  rename(State = "NAME", Total = "summary_est", 
         Non_Hispanic = "B03001_002", Hispanic = "B03001_003") |> 
  mutate(Hispanic_Pct = Hispanic / Total) |> 
  left_join(states, by = "GEOID") |> 
  st_as_sf() |> 
  shift_geometry()
```

```{r}
hisp_us <- ggplot(us_hisp_pct, aes(fill = Hispanic_Pct)) +
  geom_sf() +
  labs(title = "Hispanic Population by State",
       fill = "Percent",
       ) +
  scale_fill_viridis_c(labels = scales::percent, limits = c(0,1)) +
  theme_void() +
  theme(plot.margin = margin(.5,.5,.5,.5, "cm"))

```

:::

Let's see the top 5 states by total Latinos. Not surprisingly, the largest states have the largest population of Hispanics.

```{r}
us_hisp_pct |> 
  st_drop_geometry() |> 
  slice_max(order_by = Hispanic, n = 8) |> 
  select(State, Hispanic)
```

On the other hand, looking by percentage tells a different story:

```{r}
us_hisp_pct |> 
  st_drop_geometry() |> 
  slice_max(order_by = Hispanic_Pct, n = 8) |> 
  select(State, Hispanic_Pct)
```

As you can see, New Mexico is the state with the highest proportion of Latinos, nearly half, and almost 10% higher than the next highest states. Let's look at a map:

```{r}
#| out-width: "70%"
#| fig-align: "center"
#| echo: false

hisp_us
```


We can see how Latinos are spread across New Mexico:

::: {.callout-tip collapse=true icon=false}
## Code

```{r}
nm_hisp_pct <- get_acs(
  geography = "county",
  state = "NM",
  variables = c("B03001_002", "B03001_003"),
  summary_var = "B03001_001",
  year = 2023,
  cache_table = TRUE) |> 
  left_join(nm_counties,
            by = "GEOID") |> 
  select(-moe, -summary_moe) |> 
  pivot_wider(names_from = variable, values_from = estimate) |> 
  rename(County = "NAME", Total = "summary_est", 
         Non_Hispanic = "B03001_002", Hispanic = "B03001_003") |> 
  mutate(Hispanic_Pct = Hispanic / Total) |> 
  left_join(nm_counties, by = "GEOID") |> 
  st_as_sf()
```

```{r}
hisp_nm <- ggplot(nm_hisp_pct, aes(fill = Hispanic_Pct)) +
  geom_sf() +
  labs(title = "Hispanic Population in New Mexico",
       fill = "Percent",
       ) +
  scale_fill_viridis_c(labels = scales::percent, limits = c(0,1)) +
  theme_void()
```

:::

```{r}
#| out-width: "70%"
#| fig-align: "center"
#| echo: false

hisp_nm
```

Not only is the Hispanic population of New Mexico uniquely large in percentage terms, it is also unusual in terms of the racial identity of the Hispanics. Unlike Latinos in most of the rest of the country, many Hispanics in New Mexico do not identify at all with Mexico or Central America, but directly with their European ancestry.

The rest of this article will explore the Hispanic population in New Mexico and the historical context which explains this particular situation. Along the way, I will show how to use the R package `tidycensus` to access the wealth of data on <https://census.gov>.