---
title: Exporing New Mexico Demographics with TidyCensus
subtitle: A subtitle
description: Some description
output: 
  html_document:
    toc: true
    keep_md: true
    self_contained: true
date: 2025-05-24
code-overflow: wrap
categories: [R, TidyCensus, Data Science]
draft: false
---

```{r setup, include=FALSE, paged.print=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This article will explore some interesting aspects of New Mexico's racial demographics using R and the TidyCensus package. The US Census Bureau makes an amazing amount of data available to the public on <https://census.gov>. The site allows you to easily search for and download demographic information with myriad variables and geographic subsets. The site also provides an API for programmatic access, and R is fortunate to have the TidyCensus package to easily grab data of interest to the programmer.

New Mexico has relatively large Native American and Hispanic populations. Nearly half of New Mexico residents are Hispanic, compared to a national average of around 20% (we'll calculate it later). The profile of Latinos in New Mexico, however, is very different from the rest of the country, as we shall see.

Not only is the Hispanic population of New Mexico uniquely large in percentage terms, it is also unusual in terms of national identity of the Hispanics. Unlike Latinos in most of the rest of the country, many Hispanics in New Mexico do not identify at all with Latin America, but directly with their European ancestry.

The rest of this article will explore the Hispanic population in New Mexico and the historical context which explains this particular situation. Along the way, I will show how to use the R package `tidycensus` to access the wealth of data on <https://census.gov>.


## A little history

New Mexico has the oldest state capital in the United States, Santa Fe having been founded 10 years before Boston. While the British were exploring from the coast, the Spaniards were moving upwards from what was to be Mexico into the territory that would become New Mexico.

Mexico won independence from Spain in 1821, and that included New Mexico. While the Spanixh residents gave nominal allegiance to Mexico, it was never very enthusiastic. After the Treaty of Hidalgo, in which the territory of New Mexico was ceded to the United States, Hispanics quickly dropped any pro-Mexican sentiment, they might have had and were anxious to assert their identity as Americans with roots in Europe, not Latin America. This group is called *nuevomexicanos* by many, but in New Mexico they are called Hispanos.

## Immigrants in New Mexico

### Getting ACS data

Before looking at Latinos specifically, I would like to take a quick look at immigration to New Mexico, since many Hispanics are immigrants. Using `get_acs()` I can get data from the *American Community Survey* 5-year tables. The most time-consuming part is sifting through the thousands of variables to find the ones you want. Fortunately, `tidycensus` provides a function to download the variables, which can then be `View`ed and searched.


```{r message=FALSE}
library(tidycensus)
library(tidyverse)
library(tigris)
library(patchwork)
library(sf)
options(tigris_use_cache = TRUE)
```

```{r}
#| eval: false


v23 <- load_variables(2023, "acs5", cache = TRUE)
View(v23)
```

```{r}
pob_vars = c(
  Native = "B05002_002",
  ForeignBorn = "B05002_013"
)
```

Now I can use `get_acs()` to download the data. There are many possibilities for `geography`, which is required. Some are obvious like "us", "state" and "county". I specify which state I want. The `summary_var` argument adds an additional column to the data frame containing totals, allowing for easy calculations of percentages.

`get_acs()` has an optional argument which can be passed, `geometry = TRUE`. This handy method adds geometries from `tigris` and returns the `sf` object necessary for cartographic plotting. Unfortunately, I get errors using this for county and state data, so I need to manually perform the process.

First, I'll grab shapes for the states and for the counties in New Mexico. The `shift_geometry()` function will put Alaska and Hawaii in a convenient place on the maps.

```{r}
library(tigris)
states <- states() |> 
  select(GEOID, state = NAME, geometry) |> 
  shift_geometry()
nm_counties <- counties("NM") |> 
  select(GEOID, county = NAME, geometry)
```

Now, with standard `dplyr` syntax, I can create the `sf` data frame. I want to do comparisons to the US in general, so I will also get data for all 50 states.


```{r}
immigration_23 <- get_acs(
  geography = "county",
  state = "NM",
  variables = pob_vars,
  summary_var = "B05002_001",
  year = 2023,
  cache_table = TRUE,
) |> 
  left_join(nm_counties,  by = "GEOID") |> 
  select(variable, estimate, moe,
         county, summary_est, geometry) |> 
  st_as_sf()

```

```{r}
immigration_us <- get_acs(
  geography = "state",
  variables = pob_vars,
  summary_var = "B05002_001",
  year = 2023,
  cache_table = TRUE,
) |> 
  left_join(states,  by = "GEOID") |> 
  select(variable, estimate, moe,
         state, summary_est, geometry) |> 
  st_as_sf()
```

Percentages can be more interesting than raw numbers, so I will calculate them:

```{r}
immigration_23$pct_foreign = immigration_23$estimate / immigration_23$summary_est
immigration_us$pct_foreign = immigration_us$estimate / immigration_us$summary_est
```

Now we can see some basic stats:

```{r}
pct_fb_nm <- with(
  immigration_23, 
  mean(pct_foreign[variable == "ForeignBorn"]))
pct_fb_us <- with(
  immigration_us, 
  mean(pct_foreign[variable == "ForeignBorn"]))
data.frame(
  Area = c("New Mexico", "US"),
  Average = c(pct_fb_nm, pct_fb_us)
)
```

Despite having a the largest Hispanic population in the country, New Mexico has a much smaller percentage of immigrants than the overall average.

### Visualizing the data

Now I would like to visualize the data. (Explaining `ggplot` is beyond the scope of this article.)

```{r}
#| label: fig-fvn-col
#| out-width: "80%"
#| fig-align: "center"
#| fig-cap: Foreign vs. Native Populations

p1 <- ggplot(immigration_23, 
             aes(x = variable, y = estimate)) +
  geom_col(fill = "navy") +
  labs(x = element_blank(),
       y = "Population",
       fill = "County",
       title = "New Mexico",
       caption = "acs5 2023") +
  scale_y_continuous(labels = scales::comma) +
  theme_bw()

p2 <- ggplot(immigration_us, aes(x = variable, y = estimate)) +
  geom_col(fill = "navy") +
  labs(x = element_blank(),
       y = "Population",
       title = "US",
       caption = "acs5 2023") +
  scale_y_continuous(labels = scales::comma) +
  labs(y = element_blank()) +
  theme_bw()

p1 + p2
```

We can see on a map where the largest concentration of immigrant groups are.


```{r}
#| label: fig-imm-pop-state
#| out-width: "90%"
#| fig-align: "center"

immigration_us$percent <- immigration_us$estimate / immigration_us$summary_est

immigration_us |> 
  filter(variable == "ForeignBorn") |> 
  ggplot(aes(fill = percent)) +
  geom_sf(aes(geometry = geometry)) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(title = "Immigrant population by state",
       caption = "acs5 2023") +
  theme_void() +
  theme(plot.margin = margin(.3,.3,.3,.3, "cm"))

```


```{r}
#| label: fig-imm-pop-county
#| out-width: "90%"
#| fig-align: "center"

immigration_23 |> 
  filter(variable == "ForeignBorn") |> 
  ggplot(aes(fill = pct_foreign)) +
  geom_sf(aes(geometry = geometry)) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(title = "NM Immigrant population by county",
       caption = "acs5 2023", 
       fill = "percent") +
  theme_void() +
  theme(plot.margin = margin(.3,.3,.3,.3, "cm"))
```

Not surprisingly, the highest percentage is near the Mexican border.

### Origins of immigrants

Since we are concerned with Hispanics, it is useful to have an idea about immigration from Latin America, and Mexico in particular given that NM is a border state.

```{r}
orig_vars_23 = c(
  Europe = "B05006_002",
  Asia = "B05006_047",
  Eastern_Asia = "B05006_048",
  China = "B05006_049",
  Asia_SC = "B05006_056",
  Asia_SE = "B05006_068",
  Vietnam = "B05006_077",
  Asia_W = "B05006_079",
  Africa = "B05006_095",
  Oceania = "B05006_130",
  Latin_America = "B05006_139",
  Caribbean = "B05006_140",
  Central_America = "B05006_154",
  Mexico = "B05006_160",
  South_America = "B05006_164",
  Canada = "B05006_177"
)
```

```{r}
foreign_origin_nm <- get_acs(
  geography = "county",
  state = "NM",
  variables = orig_vars_23,
  summary_var = "B05006_001",
  year = 2023,
  cache_table = TRUE
) |> 
  left_join(nm_counties, by = "GEOID") |> 
  select(variable, estimate, moe,
            county, total = summary_est, 
            geometry) |> 
  st_as_sf()
```

```{r}
continents <- c(
  "Europe",
  "Central_America",
  "Asia",
  "South_America",
  "Africa",
  "Oceania",
  "Canada"
)
```

```{r}
#| label: fb-nm-bar
#| out-width: "90%"
#| fig-align: "center"

foreign_origin_nm |> 
  filter(variable %in% continents) |> 
  mutate(variable = fct_reorder(variable, desc(estimate))) |> 
  ggplot(aes(x = variable, y = estimate, fill = county)) +
  geom_col() +
  labs(x = "Continent",
       y = "Population",
       fill = "County",
       title = "Origins of foreign-born New Mexico population",
       subtitle = "By Continent",
       caption = "acs5 2023") +
  scale_y_continuous(labels = scales::comma) +
  theme_gray() +
  theme(axis.text.x = element_text(angle = 30),
        plot.margin = margin(.3,.3,.3,.3, "cm"))
```

The largest group of immigrants is from Central America, which includes Mexico in these tables. Zooming in on the eight most populated counties is interesting, showing a sizeable number of Asian immigrants, especially around Albuquerque (Bernalillo County).

```{r}
counties_top_8 <- foreign_origin_nm |> 
  group_by(county) |> 
  summarise(est_sum = sum(estimate)) |> 
  slice_max(order_by = est_sum, n = 8) |>
  ungroup()
```

```{r}
#| label: fig-fb-nm
#| out-width: "90%"
#| fig-align: "center"
#| fig-cap: Continents of origin for New Mexican Immigrants

foreign_origin_nm |> 
  filter(variable %in% continents & county %in% counties_top_8$county) |> 
  mutate(county = fct_reorder(county, desc(total))) |> 
  ggplot(aes(x = county, y = estimate, fill = variable)) +
  geom_col() +
  labs(x = "County",
       y = "Population",
       fill = "Continent",
       title = "Origins of foreign-born New Mexico population",
       subtitle = "Top 10 Counties by Population") +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_viridis_d(labels = c("Africa", "Asia", "Canada", "Central America",
                               "Europe", "Oceania", "South America")) +
  theme_gray() +
  theme(axis.text.x = element_text(angle = 30))
```


```{r}
#| echo: false

orig_vars_19 = c( Europe = "B05006_002", Asia = "B05006_047", Eastern_Asia = "B05006_048", China = "B05006_049", Asia_SC = "B05006_056", Asia_SE = "B05006_067", Vietnam = "B05006_076", Asia_W = "B05006_078", Africa = "B05006_091", Oceania = "B05006_117", Latin_America = "B05006_124", Caribbean = "B05006_125", Central_America = "B05006_138", Mexico = "B05006_139", South_America = "B05006_148", Canada = "B05006_161" )
```



## Hispanics

Let's look at where most Latinos live in the country.


```{r}
us_hisp_pct <- get_acs(
  geography = "state",
  variables = c("B03001_002", "B03001_003"),
  summary_var = "B03001_001",
  year = 2023,
  cache_table = TRUE
) |>
  select(-moe, -summary_moe) |> 
  pivot_wider(names_from = variable, 
              values_from = estimate) |> 
  rename(State = "NAME", 
         Total = "summary_est", 
         Non_Hispanic = "B03001_002", 
         Hispanic = "B03001_003") |> 
  mutate(Hispanic_Pct = Hispanic / Total) |> 
  left_join(states, by = "GEOID") |> 
  st_as_sf() |> 
  shift_geometry()
```


```{r}
#| label: fig-his-state-map
#| out-width: "90%"
#| fig-align: "center"

us_hisp_pct |> 
  filter(state != "Puerto Rico") |> 
  ggplot(aes(fill = Hispanic_Pct)) +
  geom_sf() +
  labs(title = "Hispanic Population by State",
       fill = "Percent",
       ) +
  scale_fill_viridis_c(labels = scales::percent, 
                       limits = c(0,.5)) +
  theme_void() +
  theme(plot.margin = margin(.5,.5,.5,.5, "cm"))
```

Let's see the top 5 states by total Latinos. Not surprisingly, the largest states have the largest population of Hispanics, and New Mexico doesn't even make the list.

```{r}
us_hisp_pct |> 
  st_drop_geometry() |> 
  slice_max(order_by = Hispanic, n = 8) |> 
  select(State, Hispanic)
```

On the other hand, looking by percentage tells a different story:

```{r}
us_hisp_pct |> 
  st_drop_geometry() |> 
  slice_max(order_by = Hispanic_Pct, n = 8) |> 
  select(State, Hispanic_Pct)
```

As you can see, New Mexico is the state with the highest proportion of Latinos, nearly half, and almost 10% higher than the next highest states.

We can see how Latinos are spread across New Mexico:

```{r}
nm_hisp_pct <- get_acs(
  geography = "county",
  state = "NM",
  variables = c("B03001_002", "B03001_003"),
  summary_var = "B03001_001",
  year = 2023,
  cache_table = TRUE) |> 
  left_join(nm_counties,
            by = "GEOID") |> 
  select(-moe, -summary_moe) |> 
  pivot_wider(names_from = variable, values_from = estimate) |> 
  rename(County = "NAME", Total = "summary_est", 
         Non_Hispanic = "B03001_002", Hispanic = "B03001_003") |> 
  mutate(Hispanic_Pct = Hispanic / Total) |> 
  st_as_sf()
```

```{r}
#| label: fig-his-pop-pct
#| out-width: "90%"
#| fig-align: "center"

ggplot(nm_hisp_pct, aes(fill = Hispanic_Pct)) +
  geom_sf() +
  labs(title = "Hispanic Population in New Mexico",
       fill = "Percent",
       ) +
  scale_fill_viridis_c(labels = scales::percent, limits = c(0,.8)) +
  theme_void() +
  theme(plot.margin = margin(.3,.3,.3,.3, "cm"))
```

The highest concentrations are not along the border, but in the Northern counties. These are the counties with the largest percentage of Hispanics, followed by the percentages in New Mexico's most populated counties.

```{r}
nm_hisp_pct |> 
  slice_max(order_by = Hispanic_Pct, n = 8) |> 
  st_drop_geometry() |> 
  select(county, Hispanic_Pct)
```

```{r}
nm_hisp_pct |> 
  filter(county %in% counties_top_8$county) |> 
  st_drop_geometry() |> 
  select(county, Hispanic_Pct)
```


# Hispanic origins

Now let's look at the national origins of Latinos in New Mexico.

```{r}
hisp_vars = c(
  Non_Hispanic = "B03001_002",
  Hispanic = "B03001_003",
  Hispanic_Mexican  = "B03001_004",
  Hispanic_PR = "B03001_005", 
  Hispanic_Cuba = "B03001_006", 
  Hispanic_DR = "B03001_007", 
  Hispanic_CA = "B03001_008", 
  Hispanic_SA = "B03001_016", 
  Hispanic_Spaniard = "B03001_028",
  Hispanic_Spanish = "B03001_029",
  Hispanic_Spanish_American = "B03001_030",
  Hispanic_White = "B03002_013"
)

hispanic_nm <- get_acs(
  geography = "county",
  state = "NM",
  variables = hisp_vars,
  summary_var = "B03001_001",
  year = 2023,
  cache_table = TRUE) |> 
  left_join(nm_counties, by = "GEOID") |> 
  select(GEOID, variable, estimate, moe,
            county, summary_est, 
            geometry) |> 
  st_as_sf()
```

```{r}
hispanic_us <- get_acs(
  geography = "state",
  variables = hisp_vars,
  summary_var = "B03001_001",
  year = 2023,
  cache_table = TRUE) |> 
  left_join(states, by = "GEOID") |> 
  select(GEOID, variable, estimate, moe,
            state, summary_est, 
            geometry) |> 
  st_as_sf()
```



```{r}
hispanic_nm <- hispanic_nm |> 
  pivot_wider(names_from = variable, values_from = estimate) |> 
  rowwise() |> 
  mutate(Hispanic_Spain = sum(c_across(Hispanic_Spaniard:Hispanic_Spanish_American), na.rm = TRUE)) |> 
  select(-c("Hispanic_Spaniard", "Hispanic_Spanish", "Hispanic_Spanish_American")) |> 
  pivot_longer(cols = Non_Hispanic:Hispanic_Spain, names_to = "variable", values_to = "estimate", values_drop_na = T) |> 
  group_by(county, variable) |> 
  mutate(pop = sum(estimate)) |> 
  distinct(county, variable, .keep_all = TRUE)
```

First, here are the proportions of Hispanic to Non-Hispanic in the biggest counties.

```{r}
#| label: fig-hvnv-bar
#| out-width: "90%"
#| fig-align: "center"

hispanic_nm |> 
  filter(county %in% counties_top_8$county & variable %in% c("Hispanic", "Non_Hispanic")) |> 
  mutate(county = fct_reorder(county, desc(estimate))) |> 
  ggplot(aes(x = county, y = estimate, fill = variable)) +
  geom_col(position = "dodge") +
  labs(x = "County",
       y = "Population",
       fill = "Continent",
       title = "Hispanic vs Non-Hispanic population in New Mexico",
       subtitle = "Top 10 Counties by Population",
       caption = "Source: census.gov acs5 2023") +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_viridis_d(labels = c("Hispanic", "Non Hispanic")) +
  theme_gray() +
  theme(axis.text.x = element_text(angle = 30)) +
  theme(plot.margin = margin(.3,.3,.3,.3, "cm"))
```

```{r}
hispanic_us <- get_acs(
  geography = "state",
  variables = hisp_vars,
  summary_var = "B03001_001",
  year = 2023,
  cache_table = TRUE) |> 
  rename(state = NAME) |> 
  pivot_wider(names_from = variable, values_from = estimate) |> 
  rowwise() |> 
  mutate(Hispanic_Spain = sum(c_across(Hispanic_Spaniard:Hispanic_Spanish_American), na.rm = TRUE)) |> 
  select(-c("Hispanic_Spaniard", "Hispanic_Spanish", "Hispanic_Spanish_American")) |> 
  pivot_longer(cols = Non_Hispanic:Hispanic_Spain, names_to = "variable", values_to = "estimate", values_drop_na = T)|> 
  group_by(state, variable) |> 
  mutate(pop = sum(estimate)) |> 
  distinct(variable, .keep_all = TRUE) |> 
  mutate(percent = pop / summary_est) |> 
  left_join(states, by = "GEOID") |> 
  select(variable, pop, summary_est,
         state = state.y, geometry) |> 
  st_as_sf()
```

And then the national identities from the same counties.

```{r}
#| label: fig-org-ident
#| out-width: "90%"
#| fig-align: "center"

hispanic_nm |> 
  filter(county %in% counties_top_8$county & !(variable %in% c("Hispanic", "Non_Hispanic"))) |> 
  mutate(county = fct_reorder(county, desc(pop))) |> 
  ggplot(aes(x = county, y = pop, fill = factor(variable, levels = c("Hispanic_White", "Hispanic_Spain", "Hispanic_Mexican", "Hispanic_CA")))) +
  geom_col(position = "dodge") +
  labs(x = "County",
       y = "Population",
       fill = "Race",
       title = "Origin Identification of Hispanic Population",
       subtitle = "Top 10 Counties by Population",
       caption = "Source: census.gov acs5 2023") +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_viridis_d(labels = c("White", "Spanish", "Mexican", "Central American")) +
  theme_gray() +
  theme(axis.text.x = element_text(angle = 30)) +
  theme(plot.margin = margin(.3,.3,.3,.3, "cm"))
```

Most Hispanics clearly identify with Mexico, as is true nationally, and White implies European, and therefore Spanish, heritage.

```{r}
#| label: fig-nat-ori
#| out-width: "90%"
#| fig-align: "center"
#| fig-cap: "National Identification of Hispanic Population in the US"

hispanic_us$percent = hispanic_us$pop / hispanic_us$summary_est
hispanic_us |> 
  filter(state != "Puerto Rico" & !(variable %in% c("Hispanic", "Non_Hispanic"))) |> 
  group_by(variable) |> 
  summarise(
    pop = sum(pop),
    percent = mean(percent)
  ) |> 
  mutate(variable = fct_reorder(variable, desc(pop))) |> 
  ggplot(aes(x = variable, y = pop)) +
  geom_col(fill = "cyan") +
  labs(x = "Nationality",
       y = "Population",
       title = "National Identification Hispanics",
       caption = "Source: census.gov acs5 2023") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_discrete(labels = c("Mexican", "White", "Central America", "Puerto Rico", 
                              "South America", "Cuba", "Dominican Republic", "Spain")) +
  theme_gray() +
  theme(
    axis.text.x = element_text(
      angle = 30, hjust = .5),
    plot.margin = margin(.3,.3,.3,.3, "cm"))
```



```{r}
#| label: fig-hisp-white
#| out-width: "90%"
#| fig-align: "center"

hispanic_us |> 
  filter(variable == "Hispanic_White") |> 
  ggplot(aes(fill = percent)) +
  geom_sf(aes(geometry = geometry)) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(title = "Hispanics with White identity") +
  theme_void() +
  theme(plot.margin = margin(.3,.3,.3,.3, "cm"))
```

New Mexico has the highest proportion of Hispanics who identify as white after Puerto Rico.

